---
title: "Mulivariate Viz"
---

Use this file for practice with the **mulivariate viz** in-class activity.
Refer to the class website for details.

```{r}
# Import and check out data
education <- read.csv("https://mac-stat.github.io/data/sat.csv")
head(education)
View(education)
```

## Exercise 1

```{r}
# Part a
# Construct a plot of how the average sat scores vary from state to state. (Just use 1 variable – sat not State!)
library(ggplot2)
ggplot(education, aes(x = sat)) + 
  geom_density()
```

#### Part b

##### Summarize your observations from the plot. Comment on the basics: range, typical outcomes, shape. (Any theories about what might explain this non-normal shape?)

They range from 800 to 1000

## Exercise 2

#### Part a

```{r}
# Construct a plot of sat vs expend
# Include a "best fit linear regression model" (HINT: method = "lm")
ggplot(education, aes(x=sat, y=expend)) +
  geom_smooth(method ="lm")
```

```{r}
# Construct a plot of sat vs salary
# Include a "best fit linear regression model" (HINT: method = "lm")
ggplot(education, aes(x=sat, y=salary)) +
  geom_smooth(method ="lm")
```

#### Part b

```{r}

```

##### What are the relationship trends between SAT scores and spending? Is there anything that surprises you?

The higher the scores the lower the salary that teachers received.

## Exercise 3

##### Construct one visualization of the relationship of sat with salary and expend. HINT: Start with just 2 variables and tweak that code to add the third variable. Try out a few things!

```{r}
ggplot(education, aes(x = sat, y = expend, color = salary)) +
  geom_point()
```

## Exercise 4

```{r}
ggplot(education, aes(y = sat, x = salary, color = cut(expend, 2))) + 
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm")
```

##### What happens if you change “2” to “3”? -

-   increases the number of lines of continuous lines of the expend variable, it also changes the color

## Exercise 5

#### Part a

Build a univariate viz of fracCat to better understand how many states fall into each category.

```{r}
ggplot(education, aes(x = fracCat)) + 
  geom_bar()
```

#### Part b

Build 2 bivariate visualizations that demonstrate the relationship between sat and fracCat.
What story does your graphic tell and why does this make contextual sense?

```{r}
ggplot(education, aes(x = sat, fill = fracCat)) + 
  geom_density(alpha = 0.5)
```

#### Part c

Make a trivariate visualization that demonstrates the relationship of sat with expend AND fracCat.
Highlight the differences in fracCat groups through color AND unique trend lines.
What story does your graphic tell?
Does it still seem that SAT scores decrease as spending increases?

```{r}
ggplot(education, aes(x = expend, y = sat, color = fracCat)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

#### Part d

Putting all of this together, explain this example of Simpson’s Paradox.
That is, why did it appear that SAT scores decrease as spending increases even though the opposite is true?

-   There is selected participation in states where there is lowere expenditure therefore produces higher results.
